Ricci-Ollivier curvature of two random walks on rooted phylogenetic subtree-prune-regraft graph

### 2015-02-13 curvature.tex
more explanation in some proofs

\begin{lemma}
\label{lem:degree_change}
Let $T$ and $S$ be phylogenetic trees such that $T$ can be obtained from $S$ by moving a subtree $R$ with $k$ leaves from its position adjacent to subtree $U$ to a location adjacent to subtree $V$.
Let $L$ be the $\lca{U,V}$ in $T$.
Let $a$ be the number of intermediate nodes on the path from the parent of $R$ to $L$ in $T$, excluding endpoints.
Similarly, let $b$ be the number of intermediate nodes on the path from $V$ to $L$ in $T$, excluding endpoints.
Let $i$ be the number of leaves in $U$, excluding any leaves of $R$, and $j$ be the number of leaves in $V$, excluding any leaves of $R$. \todo{My impression was that the leaves of $R$ and $U$ (or $V$) were disjoint. If that's the case, then this "excluding" clause is more confusing than helpful.}
Then the degrees of $T$ and $S$ differ by:
$$2\left(k(a-b) + i - j\right).$$

@@

	If the subtree of $k$ leaves is moved to an ancestor, then we gain an additional leaf\todo{Hm, I'm not sure where this additional leaf is!} to balance the product but increase $j$.
	This corresponds to moving the bottom subtree of $\floor{\frac{n}{2}}$ or $\ceil{\frac{n}{2}}$ leaves in a ladder tree to the root.
	Namely, we have $2(ka + 1 - j)$, where $j = n - k = a + 1$.
	If we move the additional leaf, we have:
$$\degree{S} - \degree{T} \le 2\left(\ceilXXL{\frac{n}{2}}\floorV{\frac{n-2}{2}}  + 1 - \left(\floorV{\frac{n-2}{2}} + 1\right)\right) = 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}}.$$
Similarly, if we do not move the additional leaf, we have:
$$\degree{S} - \degree{T} \le 2 \left(\ceilV{\frac{n-2}{2}}\floorXXL{\frac{n}{2}} +1 -  \left(\ceilV{\frac{n-2}{2}} + 1\right)\right) = 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}}.$$

	The relative change in degree, $\frac{\degree{T}}{\degree{S}}$, is minimized when $\degree{T}$ is as small as possible and $\degree{S} - \degree{T}$ is maximal.\todo{This sort of thing does not appear true in general. Please explain why it works here.}
	That is, $\degree{S} = 3n^2 - 13n + 14$ and $\degree{S} + 2\floor{\frac{n-2}{2}} \ceil{\frac{n-2}{2}}$.
	We then have:
	\begin{align*}
		\frac{\degree{T}}{\degree{S}} &= \frac{3n^2 - 13n + 14}{3n^2 - 13n + 14 + 2\floor{\frac{n-2}{2}} \ceil{\frac{n-2}{2}}} \\
		&\ge \frac{3n^2 - 13n + 14}{3n^2 - 13n + 14 + \frac{1}{2}(n-2)^2} \\
		&\ge \frac{3n^2 - 13n + 14}{3.5n^2 - 15n + 16}
	\end{align*}
	Thus (2) and (3) follow.

>>>>

\begin{lemma}
\label{lem:degree_change}
Let $T$ and $S$ be phylogenetic trees such that $T$ can be obtained from $S$ by moving a subtree $R$ with $k$ leaves from its position adjacent to subtree $U$ to a location adjacent to subtree $V$.
Let $L$ be the $\lca{U,V}$ in $T$.
Let $a$ be the number of intermediate nodes on the path from the parent of $R$ to $L$ in $T$, excluding endpoints.
Similarly, let $b$ be the number of intermediate nodes on the path from $V$ to $L$ in $T$, excluding endpoints.
Let $i$ be the number of leaves in $U$ and $j$ be the number of leaves in $V$, excluding any leaves of $R$. \todo{EM: if we move R to an ancestor (e.g. the root) then the subtree $V$ in $T$ contains both $R$ and $U$. $R$ and $U$ are disjoint in $T$, though so I dropped one excluding.}
Then the degrees of $T$ and $S$ differ by:
$$2\left(k(a-b) + i - j\right).$$

@@

	If the subtree of $k$ leaves is moved to an ancestor, then we do not need to exclude the target branch from $k$ and $a$, gaining an additional leaf \todo{Hm, I'm not sure where this additional leaf is! EM: does this make more sense?} to balance the product at the cost of increasing $j$.
	This corresponds to moving the bottom subtree of $\floor{\frac{n}{2}}$ or $\ceil{\frac{n}{2}}$ leaves in a ladder tree to the root.
	Namely, we have $2(ka + 1 - j)$, where $j = n - k = a + 1$.
	If we move the additional leaf, we have:
$$\degree{S} - \degree{T} \le 2\left(\ceilXXL{\frac{n}{2}}\floorV{\frac{n-2}{2}}  + 1 - \left(\floorV{\frac{n-2}{2}} + 1\right)\right) = 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}}.$$
Similarly, if we do not move the additional leaf, we have:
$$\degree{S} - \degree{T} \le 2 \left(\ceilV{\frac{n-2}{2}}\floorXXL{\frac{n}{2}} +1 -  \left(\ceilV{\frac{n-2}{2}} + 1\right)\right) = 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}}.$$

The relative change in degree, $\frac{\degree{T}}{\degree{S}}$, can also be written as $\frac{\degree{T}}{\degree{T} + (\degree{S} - \degree{T})$}.
This is is minimized when $\degree{T}$ is as small as possible and $\degree{S} - \degree{T}$ is maximum. \todo{This sort of thing does not appear true in general. Please explain why it works here. EM: Is this better? Think of $\frac{x}{1+x}$ which is an increasing function.}
	That is, $\degree{T} = 3n^2 - 13n + 14$ and $\degree{S} = \degree{T} + 2\floor{\frac{n-2}{2}} \ceil{\frac{n-2}{2}}$.
	We then have:
	\begin{align*}
		\frac{\degree{T}}{\degree{S}} &= \frac{3n^2 - 13n + 14}{3n^2 - 13n + 14 + 2\floor{\frac{n-2}{2}} \ceil{\frac{n-2}{2}}} \\
		&\ge \frac{3n^2 - 13n + 14}{3n^2 - 13n + 14 + \frac{1}{2}(n-2)^2} \\
		&\ge \frac{3n^2 - 13n + 14}{3.5n^2 - 15n + 16}
	\end{align*}
	Thus (2) and (3) follow.
